---
title: "Spatialize data"
author: "Anna Talucci"
date: "2026-05-13"
output: html_document
---

```{r clear environment, include=FALSE}
rm(list=ls())
```

# Overview

**PROJECT**
FCS-FireAlt project

**PURPOSE OF SCRIPT**
Spatialize FIRE ALT dataset

**INPUTS**
- Data downloaded from Arctic Data Center https://arcticdata.io/catalog/view/doi%3A10.18739%2FA2RN3092P

**OUTPUTS**

**CRITICAL DETAILS**

**WORKFLOW SUMMARY**
1. 


# Packages
```{r}
library(tidyverse)
library(sf)
library(mapview)
```


# Data

```{r}
pair_df = read_csv("../data/fire_alt/FireALTEstimatedPairsBurnedUnburned.csv")
raw_df = read_csv("../data/fire_alt/FireAltEstimatedRawData.csv")
```

# View Data

```{r}
pair_df
raw_df
```

# Summarise raw based on pair and lat lon

```{r}
( alt_df = raw_df %>% 
  group_by(submitNm, distur, paired, fireYr, year, tsf, tsfClass, permaExtent, lat, lon) %>%
  summarise(
    cntryId = paste(unique(cntryId), collapse = ', '),
    #lastNm = paste(unique(lastNm), collapse = ', '),
    #lat = mean(lat),
    #lon = mean(lon),
    biome = paste(unique(biome), collapse = ', '),
    #resName = paste(unique(resName), collapse = ', '), 
    #resBiomeName = paste(unique(resBiomeName), collapse = ', '), 
    #fireId = paste(unique(fireId), collapse = ', '),
    #orgDepth = mean(orgDepth),
    #srfH2O = paste(unique(srfH2O), collapse = ', '),
    #topoPos = paste(unique(topoPos), collapse = ', '),
    slope = mean(slope),
    #vegCvr = paste(unique(vegCvr), collapse = ', '),
    tsf = paste(unique(tsf), collapse = ', '),
    tsfClass = paste(unique(tsfClass), collapse = ', '),
    #estDoy = mean(estDoy),
    estDepth = mean(estDepth),
    n = n() ) %>%
    ungroup()
)
```

```{r}
glimpse(alt_df)
```

```{r}
sort(unique(alt_df$fireYr))
sort(unique(alt_df$year))
```
## summary tables

```{r}
(burned_summary = alt_df %>% 
  filter(distur == 'burned') %>%
  group_by(fireYr, year, permaExtent, cntryId, biome) %>%
  summarize(n = n()) %>%
  arrange(fireYr)
)
  
```

```{r}
write_csv(burned_summary, "../outputs/summaries/2026-05-22_FireALT_burned_summary.csv")
```

```{r}
(summary = alt_df %>% 
   filter(distur == "burned") %>%
  filter(fireYr <= 1985) %>%
  group_by(fireYr, year, permaExtent, cntryId, biome) %>%
  summarize(n = n()) %>%
  arrange(fireYr)
)
  
```

# Create points shapefile

## Function to make points

```{r}
df_to_sf <- function(x){
  st_as_sf(x, coords = c("lon","lat"), crs = 4326, remove = FALSE)
}
```

## Drop Lat Lon NA and convert to shapefile 

```{r}
( alt_pts = alt_df %>% 
    df_to_sf() )
```

# plot

```{r}
my_colors <- c("#E41A1C", "#377EB8") 

mapview(alt_pts, zcol = "distur", col.regions = my_colors)
```


# Write to shapefile

```{r eval=FALSE, include=FALSE}
st_write(alt_pts, "../outputs/shapefile/2026-05-13_FireAltPoints.shp", driver="ESRI Shapefile")
```


